Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2021-02-24 ManuscriptTime.tex
EOD

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies, translating into a significant
reduction in computation time. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after which the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equations in an isotropic, homogeneous medium is given by \citep[e.g.,][
equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the buffer zone outside of the
area where source and receivers are located.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes. It is
important to note that while we show the transform from frequency to time
domain, the conclusions regarding the Fourier transform can equally well be
applied to the transform from time-domain to frequency-domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \citep{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, private
communication, 2020).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equations, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
20.04 and Python 3.8. All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent. For the semi-analytical computations of layered models we use the
open-source (Apache License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

@@

\rtodo \added{Mention again that important for the transform are accurate
results; these gridding as what we implemented in the 3D code.}

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the perfectly
electrically conducting boundary condition. It can also have a much higher
stretching factor $\alpha_\mr{c}$. In the presented examples we have chosen
$D_\mr{c}$ such that the distance for the signal diffusing from the source to
the boundary and back to the receiver closest to the boundary is at least two
wavelengths, after which the amplitude of the signal is reduced to a millionth
of its initial strength. The wavelength $\lambda$ to compute $D_\mr{c}$ is
given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth. Note that
for shallow marine and land cases this also applies to the horizontal
dimensions, not only to the upward $z$-direction and similarly for very
resistive basements, even in deep water. One way to circumvent this difficulty
is the use of a primary-secondary formulation, where the primary field,
including the air wave, is computed with a semi-analytical code for layered
media. We do not consider this approach here.

@@

>>>>

\rtodo \added{Extra example: No-IP vs IP; land, 1D}

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies \added{or times}, translating into
a significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required\replaced{, which}{. We build our approach upon
\mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain} is naturally based on the skin depth,
\deleted{which is} the distance after which the amplitude of the
electromagnetic field is decayed by $1/e\approx 37\,\%$. The skin depth
$\delta$ is a function of conductivity and frequency, and for the diffusive
approximation of Maxwell's equations in an isotropic, homogeneous medium is
given by \citep[e.g.,][ equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\added{We build our approach upon} \cite{GEO.07.Plessix}\added{, who presented
such an adaptive gridding for multi-frequency (and multi-source) CSEM
modelling. They} define the minimum cell width $\Delta_\mathrm{min}$ as
a fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the buffer zone outside of the
area where source and receivers are located.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes. It is
important to note that while we show the transform from frequency to time
domain, the conclusions regarding the Fourier transform can equally well be
applied to the transform from time-domain to frequency-domain data for a code
that solves Maxwell's equations in the time domain. \deleted{A variation of our
method was implemented, for instance, by \custem \mbox{\citep{GEO.19.Rochlitz}}
to obtain frequency-domain data from time-domain computations (Rochlitz,
private communication, 2020).}

@@

\added{The requirement for any transform is a robust solver of sufficient
precision.} \replaced{We use f}{F}or the numerical computations \deleted{we
use} the open-source (Apache License 2.0) code\added{s \empymod
\mbox{\citep{GEO.17.Werthmuller}} and} \emg3d
\citep{JOSS.19.Werthmuller}\replaced{.}{,} \added{The former computes
semi-analytical responses of layered models, the latter is} a multigrid solver
based on \cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov
subspace solvers or as a solver on its own. The multigrid approach works fine
for the diffusive approximation of Maxwell's equations, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
20.04 and Python 3.8. All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent. \deleted{For the semi-analytical computations of layered models we
use the open-source (Apache License 2.0) code \empymod
\mbox{\citep{GEO.17.Werthmuller}}.}

@@

\added{The proposed Fourier transformation requires a robust solver that can
compute precise results over a wide range of frequencies (times) to obtain
time-domain (frequency-domain) responses. In our case the solver uses a
stretched, regular grid and computes the electric fields in the frequency
domain. This setup is the target of the following gridding recommendations.
Whilst they will look differently for other mesh types or a time-domain code,
some conclusion will still hold.}

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the perfectly
electrically conducting boundary condition. It can also have a much higher
stretching factor $\alpha_\mr{c}$. In the presented examples we have chosen
$D_\mr{c}$ such that the distance for the signal diffusing from the source to
the boundary and back to the receiver closest to the boundary is at least two
wavelengths, after which the amplitude of the signal is reduced to a millionth
of its initial strength. The wavelength $\lambda$ to compute $D_\mr{c}$ is
given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth. Note that
for shallow marine and land cases this also applies to the horizontal
dimensions, not only to the upward $z$-direction and similarly for very
resistive basements, even in deep water. One way to circumvent this difficulty
is the use of a primary-secondary formulation, where the primary field,
including the air wave, is computed with a semi-analytical code for layered
media. We do not consider this approach here. \added{Note that grid stretching
for complex-valued diffusion fields is essentially what is done for wave fields
with the perfectly matched layer (PML). PML is an absorbing layer to avoid
scattering from the boundary by letting the field die out to zero, which is
achieved by introducing a complex factor which causes damping. As
electromagnetic diffusion fields are damped fields by themselves it suffices to
stretch the grid.}

@@

\added{mention independence of spatial complexity, why we also show 1D
examples.}
